Characterization of gaps in directly bonded Si compound optics

### 2014-04-21 AO_bondedSi.tex
tweaks from DTJ edits, add table for meshes

While it is possible that proper precautions to eliminate all micron-sized particles from both Si surfaces prior to bonding, the power-law nature of the particle size distribution in clean-room environments \cite{doi:10.1080/02786828608959094} means that it is significantly more difficult to assure that no particles of the 10$-$60 nm size scale are present.  Small particles could be a concern if they act as "tent poles", that create gaps where the bonding is incomplete (Figure \ref{figParticle}).  Authors have referred to these large ($\sim$ 1 $\mu$m size) gaps as \emph{IR bubbles}, or simply bubbles, owing to their appearance in infrared (IR) transmission images as contiguous regions of a deficit of IR transmission.  They have also been referred to as voids, or interfacial voids, or gaps.  In this paper we will reserve the term ``IR bubbles'' for gaps that are large enough to be detected directly in conventional IR imaging.  The term ``gap'' will apply more broadly to any finite gap, which exceeds the scale over which interfacial molecular forces are acting, roughly say 2 nm.

@@

The three important characteristics of the gaps are their axial extent, their lateral areal extent, and their areal fill factor.  We bored holes in silicon with inductively coupled plasma etching.  We used two different gases with low and high etch rates.  CHF$_3$ on Si exhibited about 0.3 nm/second etch rate, SF$_6$ exhibited about 1 micron/minute.  The area exposed to the process gases was about 1 inch diameter circles, or 15 mm $\times$ 20 mm rectangles.  We protected the unexposed areas with thick photoresist.  We achieved etch depths of 15$\pm$5 nm and 4000$\pm$100 nm, as measured with Veeco NT9100 Optical Profiler for the small depth, and Dektak stylus profilometry for the large depth.  Table \ref{tabbondexper} has more information about the bonding samples.  Figure \ref{figVG03pattern} shows a schematic of VG03 and its pattern, as described in Table \ref{tabbondexper}.  In a second round of experiments we achieved gap sizes of 14$-$95 nm on four substrates with three meshes with coarse, medium, and fine boxes, each with 50\% fill factors.  For measurements taken with the Veeco NT9100 Optical Profiler, uncertainties were constructed by inspecting the histogram of topology, as shown in Figure \ref{figVS20pattern}.  The large-scale distortions were removed from the topology by masking and flattening to ensure that the uncertainties accurately reflect the distribution of measured heights.  The stylus profiliometry measured values were simply assigned an uncertainty of 5\%, which was consistent with our intuition and experience.  Table \ref{tabbondexper} lists part numbers and descriptions for our bonded substrates.

@@

\begin{table}[h!]
\caption{UTexas Si bonding experiments \label{tabbondexper}}
\begin{center}
    \begin{tabular}{ c c c c c c c}
    \hline
    Name & Bonded/DSP & Diameter & Thickness & Pattern Type & Pattern Depth & Notes \\ 
    - & - & mm & mm & - & nm & - \\ 
        \hline
    VG01 & Bonded  & 100  &$1$ $\times$ 2 & None  & - &  \\
    VG02 & Bonded & 100 & $3.3$ $\times$ 2 &  None  & - &  \\
    VG03 & Bonded & 100 & $3.3$ $\times$ 2 &  Hole \&Petal & 4000$\pm$200 & See Fig. \ref{figVG03pattern}\\    
    VG04 & Bonded & 100 & $1$ $\times$ 2 &  Patchwork Islands & 15 & \\
    VG05 & DSP & 100 & 1&  - & - & Reference Sample  \\
    VG06 & DSP & 100 & 3.5 &  - & - & Reference Sample \\
    VG07 & - & - & - &  - & - & \\
    VG08 & - & - & - &  - & - & \\
    VG09 & Bonded & 75 & 1.5? & mesh C & 49 $\pm$6 & \\
    VG15 & Bonded & 100 & 1.5? &  mesh C, M, F  &14 $\pm$2 & \\
    VG17 & Bonded & 75 & 6.0? &  mesh C & 64 $\pm$6 nm & \\
    VS20 & Bonded & 100 & 1.5? &  mesh C, M, F & 95 $\pm$5 & \\
    \hline
    \end{tabular}
\end{center}
\end{table}

\section{Substrate preparation}
We cleaned the surfaces before bonding to minimize the interfacial particle density.  The surface roughness was typically about 2 nm, as measured with a Veeco NT9100 Optical Profiler.  We measured the large scale surface flatness of the pucks.  The Fisba2 interferometer has a 2 inch diameter beam at $\lambda=$632.8 nm.  We found a typical peak to valley surface flatness of 2 waves over the central 2 inch diameter.  We prepared the substrates with standard cleaning procedures of solvents in a megasonic.  We then applied MHz frequency oxygen plasma ashing.  We soaked the wafers in DI water then dried with an N$_2$ gun.  We pressed the wafers together from the center to the outside.

@@

which has an average value of about 53\%.  Note that this transmission is above a na\"ive value of $T_{DSP}=(1-R^2)$, which does not take into account multiple incoherent reflections, and is therefore an underestimate.  If the Si wafer path is less than the coherence length for the given spectral bandwidth, then the multiple reflections would interfere coherently, and the Si wafer would behave as a Fabry-P\`erot etalon.  This scenario is beyond the scope of this paper, since we are primarily interested in thick substrates with small gaps in between the interfaces.
%We work out this behavior in the Appendix and show that in the limit $l>> \lambda^2 / \delta \lambda $ the spectral transmission tends to equation X.

@@

We also looked for the indication of large gaps detectable as ``IR bubbles'' \cite{1992JEMat..21..669M} with the coarse IR imaging technique.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic white light illumination source.  We estimate that the lamp-detector combination was sensitive to wavelengths in the range 1.1-1.6 $\mu$m.  The field of view sub-sampled the 4 inch diameter wafers, so we dithered the sample.  No effort was made to flat field the detector, so copious detector non-uniformities and vignetting are present in our images.  It was easy to detect IR bubbles despite the artifacts since the bubbles translate with the wafer during dithering.  We detected IR bubbles in all 3 of the bonded Si pucks on which we recorded IR images.  The bubble areal density varied from about 4 per 100 mm diameter bonded wafer pair to over 20 per bonded wafer pair.  Naturally, the unbonded DSP wafers show no such IR bubbles, which are entirely attributable to the gaps between bonded substrates as discussed in the literature (see Section \ref{secHistory}).  Figure \ref{IRbubble} shows an example of one of the IR bubbles in which several bright and dark fringes are clearly discernible.  No attempt at quantitative measurement was made on the IR images due to evidence that the detector is non-linear.  The morphology was clear regardless of the detector response.

@@

\subsection{Constructing the predicted transmission}
We constructed the predicted transmission in the following way.  If IR imaging was unavailable, our optimistic prior belief was simply that the bond should be perfect and indistinguishable from a solitary DSP Si wafer, (i.e. 100\% relative transmission).  For the cases where we had bored gaps of known dimension and fill factors we could predict the transmission with the Fabry-Perot machinery described in section ref{secTheory}, specifically equation \ref{eqFP}.  For non-unit fill factor, we simply average 100\% transmission with the predicted gap transmission, weighted by the estimated fill factor.  Notably the meshes we designed were all 50\% fill factor.  No attempt was made to verify the delivered fill factor, but since high precision lithographic processes were employed, we take the design fill factor as gospel.  The depth of the meshes has an associated uncertainty listed in \ref{tabbondexper}.  We incorporated the uncertainty in the mesh depth in the following way.  We computed the transmission curve for the 1$\sigma$ lower and upper bounds, and plot a filled band marked as the prediction band.  The uncertainties in the mesh depths were low enough that the band spread includes all intermediate lines of predicted transmission between plus and minus 1$\sigma$.  In cases where the IR imaging was unavailable the mesh position was only coarsely known, so our prediction band had to incorporate our uncertainty about whether or not our measurement beam was on or off or overlapping the mesh border.  If the beam was off the mesh we assumed 100\% relative transmission, so we simply set the lower bound of the predicted transmission to the equivalent of a zero gap.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=.8\columnwidth]{figs/20130911_VS21posF1}}
\caption{IR transmission spectrum of part VS21 (F1) showing evidence for an excess gap\label{VS21specf1}.  Example spectrum through part VS21 at a position through the mesh gap of known dimensions.}
\end{figure}

\begin{table}[h!]
  \caption{Interpretation of the diversity of IR spectra}
  \begin{center}
    \begin{tabular}{ |p{8cm}| p{8cm} | p{2.2cm} |}
    \hline
    Appearance of spectra & Interpretation & Exemplars \\
    \hline
    Measurement within the prediction band & Life is good & VS21-M1, Figure \ref{VS21specm1}, VG15, VG07\\
    \hline
    Measured spectrum below the prediction band but shape is similar & Gap was slightly larger than predicted or bonding was incomplete& VS21-F1, Figure \ref{VS21specf1}, VS09, VG15, \\
    \hline
    Measured spectrum above the prediction band but shape is similar & Gap was slightly smaller than predicted & VS21-M2\\
    \hline
    Measured spectrum exceeds DSP reference puck & unanticipated anti-reflection coating on the surface of the Si sample (e.g. Sharpie marker residue) & VG08\\
    \hline
    High frequency, low amplitude wiggles superimposed upon smooth underlying trend & Large, unanticipated gap, with a small fill factor  & VS21-C3\\   
     																			     & wavelength dependent diffraction from the mesh patterns & VS21-C3\\       
     																			     & measurement artifact  & VS21-C3\\       												
    \hline
    Non-unit transmission constant with wavelength (gray spectrum) & Artifact of measurement- vignetting of beam, absorptive coating, zero-point drift & VS21-C4 \\
    \hline
    Completely different spectrum than anticipated with low absolute transmission & Large, inhomogeneous gap consistent with an IR bubble & VG09 \\
    \hline
    \end{tabular}
  \end{center}
\end{table}

@@

\subsection{Mitigation of measurement artifacts}
The principle artifacts and their interpretation are listed in table \ref{1}.  We mitigated the artifacts in the following ways.  For the Sharpie marker behaving as an anti-reflection (AR) coating, we simply removed the sharpie with IPA.  We have mostly been able to rule out wavelength-dependent diffraction from the mesh pattern as a measurement artifact, since the beam size is large and the angular spread of diffraction from the mesh pattern would be much smaller than the spread of the beam.  Still, to remove this concern, we moved to using the coarse mesh rather than the medium or fine meshes, when possible.  We saw no difference in the bond quality surrounding the medium or fine meshes, which is consistent with results from the literature (see the work of Mitani and collaborators from the mid 1990s).  Namely, Mitani showed that the gas pressure of bonded wafers grew as the fill factor of voids decreased.  In other words, what matters is the fill factor of voids, not their relative sizes.  Since the fine, medium, and coarse meshes had the same fill factors, we anticipate no difference in the pressure of the voids, and therefore no difference in the bonding properties.

@@

The consequence of our immersion grating design is a non-normal incidence interface in double pass.  In this case, the Fabry-P\'{e}rot framework developed here fails to predict the spectrum of the interface loss attributable to a finite albeit small (few nm) gap.  Specifically, complete internal reflection and evanescent waves will matter at large incidence angles, and the incoherent matrix technique is no longer accurate.  Polarization effects are important.  We do not have a mechanism for evaluating the bond performance until the final device is completed and its wavelength-dependent transmission is measured.  We can detect the performance degradation on final devices in the following nuanced way.  We will measure the efficiency of the bonded Si immersion grating as a function of both wavelength and angle.  Our group built a custom tool to make these measurements- the Custom Robotic Order, Wavelength, and Blaze Angle Recorder (CROWBAR) \cite{2012SPIE.8450E..2SG}.  Figure 3 in \cite{2012SPIE.8450E..2SG} shows the directly measured spectrum of a monolithic immersion grating.  Notably, the spectrum in Figure 3 in \cite{2012SPIE.8450E..2SG}  shows dozens of blaze peaks where the efficiency should be maximized.  A perfectly contacted Si$-$Si direct bonded optic would have identical performance to a monolithic counterpart.  Bonded immersion gratings with finite gap sizes will superimpose a wavelength dependent transmission loss on top of the blaze peaks.  In this way, the presence or absence of deleterious gaps can be identified in final immersion gratings.  One drawback of this approach is that gaps can only be identified after the costly process of producing the final optical device.  Large gaps identified this late in the game could render the costly optic useless, both in terms of efficiency and wavefront performance.  Our group is exploring the effectiveness of high-temperature annealing to eliminate IR bubbles in completed devices.

>>>>

While it is possible that proper precautions to eliminate all micron-sized particles from both Si surfaces prior to bonding, the power-law nature of the particle size distribution in clean-room environments \cite{doi:10.1080/02786828608959094} means that it is significantly more difficult to assure that no particles of the 10$-$60 nm size scale are present.  Small particles could be a concern if they act as "tent poles", that create gaps where the bonding is incomplete (Figure \ref{figParticle}).  Some authors \cite{1991JaJAP..30..615M, 1992JEMat..21..669M, 1989JaJAP..28L2141L, Mitani1990} have referred to these large ($\sim$ 1 $\mu$m size) gaps as \emph{IR bubbles}, or simply bubbles, owing to their appearance in infrared (IR) transmission images as contiguous regions of a deficit of IR transmission.  The gaps have also been referred to as voids\cite{2000RScI...71.1869G}, interfacial voids, or gaps.  In this paper we will reserve the term ``IR bubbles'' for gaps that are large enough to be detected directly in conventional IR imaging.  The term ``gap'' will apply more broadly to any finite size gap, exceeding the scale over which interfacial molecular forces are acting, roughly $>$2 nm.

@@

The three important characteristics of the gaps are their axial extent, their lateral areal extent, and their areal fill factor.  We bored holes in silicon with inductively coupled plasma etching.  We used two different gases with low and high etch rates.  CHF$_3$ on Si exhibited about 0.3 nm/second etch rate, SF$_6$ exhibited about 1 micron/minute.  The area exposed to the process gases was about 1 inch diameter circles, or 15 mm $\times$ 20 mm rectangles.  We protected the unexposed areas with thick photoresist.  We achieved etch depths of 15$\pm$5 nm and 4000$\pm$100 nm, as measured with Veeco NT9100 Optical Profiler for the small depth, and Dektak stylus profilometry for the large depth.  Table \ref{tabbondexper} has more information about the bonding samples, and Table \ref{tabPatternFills} lists the properties of the patterns.  The fill factor is defined as the pattern area covered by gaps relative to the total pattern area.  Figure \ref{figVG03pattern} shows a schematic of VG03 and its pattern, as described in Table \ref{tabbondexper}.  In a second round of experiments we achieved gap sizes of 14$-$95 nm on four substrates with three meshes with coarse, medium, and fine boxes, each with 50\% fill factors.  For measurements taken with the Veeco NT9100 Optical Profiler, uncertainties were constructed by inspecting the histogram of topology, as shown in Figure \ref{figVS20pattern}.  The large-scale distortions were removed from the topology by masking and flattening to ensure that the uncertainties accurately reflect the distribution of measured heights.  The stylus profiliometry measured values were simply assigned an uncertainty of 5\%, which was consistent with our intuition and experience.  Table \ref{tabbondexper} lists part numbers and descriptions for our bonded substrates.

@@

\begin{table}[h!]
\caption{UTexas Si bonding experiments \label{tabbondexper}}
\begin{center}
    \begin{tabular}{ c c c c c c c}
    \hline
    Name & Bonded/DSP & Diameter & Thickness & Pattern Type & Pattern Depth & Notes \\ 
    - & - & mm & mm & - & nm & - \\ 
        \hline
    VG01 & Bonded  & 100  &$1$ $\times$ 2 & None  & - &  \\
    VG02 & Bonded & 100 & $3.3$ $\times$ 2 &  None  & - &  \\
    VG03 & Bonded & 100 & $3.3$ $\times$ 2 &  Hole \&Petal & 4000$\pm$200 & See Fig. \ref{figVG03pattern}\\    
    VG04 & Bonded & 100 & $1$ $\times$ 2 &  Patchwork Islands & 15 & \\
    VG05 & DSP & 100 & 1&  - & - & Reference Sample  \\
    VG06 & DSP & 100 & 3.5 &  - & - & Reference Sample \\
    VG09 & Bonded & 75 & 1.5? & mesh C & 49 $\pm$6 & \\
    VG15 & Bonded & 100 & 1.5? &  mesh C, M, F  &14 $\pm$2 & \\
    VG17 & Bonded & 75 & 6.0? &  mesh C & 64 $\pm$6 nm & \\
    VS20 & Bonded & 100 & 1.5? &  mesh C, M, F & 95 $\pm$5 & See Fig. \ref{figVS20pattern}\\
    \hline
    \end{tabular}
\end{center}
\end{table}

\begin{table}[h!]
\caption{Patterned gap properties \label{tabPatternFills}}
\begin{center}
    \begin{tabular}{ c c c c }
    \hline
    Pattern name & fill factor & minimum feature size & bulk description \\ 
    - & \% & $\mu$m & '- \\ 
    Hole & 100 & - & 25 mm diameter hole \\     
    Petal & $\sim1$ & $\sim$2000 & long petal lines, Fig. \ref{figVG03pattern} \\         
    Mesh F & 50 & 40 &  100 $\mu$ square holes, Fig. \ref{figVS20pattern}\\ 
    Mesh M & 50 & 200 &  500 $\mu$ square holes\\ 
    Mesh C & 50 & 620 &  1500 $\mu$ square holes\\     
        \hline
    \end{tabular}
\end{center}
\end{table}

\section{Substrate preparation}
We cleaned the surfaces before bonding to minimize the interfacial particle density.  The surface roughness was typically about 2 nm, as measured with a Veeco NT9100 Optical Profiler.  We measured the large scale surface flatness of the pucks.  The Fisba2 interferometer has a 2 inch diameter beam at $\lambda=$632.8 nm.  We found a typical peak to valley surface flatness of 2 waves over the central 2 inch diameter.  We prepared the substrates with standard cleaning procedures of solvents in a megasonic.  We then applied MHz frequency oxygen plasma ashing.  We soaked the wafers in DI water then dried with an N$_2$ gun.  We pressed the Si substrates together from the center to the outside.

@@

which has an average value of about 53\%.  Note that this transmission is above a na\"ive value of $T_{DSP}=(1-R^2)$, which does not take into account multiple incoherent reflections, and is therefore an underestimate.  If the Si wafer path thickness is less than the coherence length for the given spectral bandwidth, then the multiple reflections would interfere coherently, and the Si wafer would behave as a Fabry-P\`erot etalon.  This scenario is beyond the scope of this paper, since we are primarily interested in thick substrates with small gaps in between the interfaces.
%We work out this behavior in the Appendix and show that in the limit $l>> \lambda^2 / \delta \lambda $ the spectral transmission tends to equation X.

@@

We looked for the indication of large gaps detectable as ``IR bubbles'' \cite{1992JEMat..21..669M} with the coarse IR imaging technique.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic white light illumination source.  We estimate that the lamp-detector combination was sensitive to wavelengths in the range 1.1-1.6 $\mu$m.  The field of view sub-sampled the 4 inch diameter wafers, so we dithered the sample.  No effort was made to flat field the detector, so copious detector non-uniformities and vignetting are present in our images.  It was easy to detect IR bubbles despite the artifacts since the bubbles translate with the wafer during dithering.  We detected IR bubbles in all 3 of the bonded Si pucks on which we recorded IR images.  The bubble areal density varied from about 4 per 100 mm diameter bonded wafer pair to over 20 per bonded wafer pair.  Naturally, the unbonded DSP wafers show no such IR bubbles, which are entirely attributable to the gaps between bonded substrates as discussed in the literature (see Section \ref{secHistory}).  Figure \ref{IRbubble} shows an example of one of the IR bubbles in which several bright and dark fringes are clearly discernible.  No attempt at quantitative measurement was made on the IR images due to evidence that the detector is non-linear.  The morphology was clear regardless of the detector response.

@@

\subsection{Constructing the predicted transmission}
We constructed the predicted transmission in the following way.  If IR imaging was unavailable, our optimistic prior belief was simply that the bond should be perfect and indistinguishable from a solitary DSP Si wafer, (i.e. 100\% relative transmission).  For the cases where we had bored gaps of known dimension and fill factors we could predict the transmission with the Fabry-Perot machinery described in section \ref{secTheory}.  Specifically we apply Equation \ref{eqFP}.  For non-unit fill factor, we simply average 100\% transmission with the predicted gap transmission, weighted by the estimated fill factor listed in Table \ref{tabPatternFills}.  No attempt was made to verify the delivered fill factor, but since high precision lithographic processes were employed, we assume the delivered fill factor is equal to the designed fill factor with no uncertainty.  The depth of the meshes has an associated uncertainty listed in \ref{tabbondexper}.  We incorporated the uncertainty in the mesh depth in the following way.  We computed the transmission curve for the 1$\sigma$ lower and upper bounds, and plot a filled band marked as the prediction band.  The uncertainties in the mesh depths were low enough that the band spread includes all intermediate lines of predicted transmission between plus and minus 1$\sigma$.  In cases where the IR imaging was unavailable the mesh position was only coarsely known, so our prediction band had to incorporate our uncertainty about whether or not our measurement beam was on or off or overlapping the mesh border.  If the beam was off the mesh we assumed 100\% relative transmission, so we simply set the lower bound of the predicted transmission to the equivalent of a zero gap.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=.8\columnwidth]{figs/20130911_VS21posF1}}
\caption{IR transmission spectrum of part VS21 (F1) showing evidence for an excess gap\label{VS21specf1}.  The spectrum was taken at a position through a mesh gap of depth 95 $\pm$5 with 50\% fill factor.  The predicted spectrum lines up with the measurement to within the measurement and model uncertainties.}
\end{figure}

\begin{table}[h!]
  \caption{Interpretation of the diversity of IR spectra \label{tabInterpretation}}
  \begin{center}
    \begin{tabular}{ |p{8cm}| p{8cm} | p{2.2cm} |}
    \hline
    Appearance of spectra & Interpretation & Exemplars \\
    \hline
    Measurement within the prediction band & Life is good & VS21-M1, Figure \ref{VS21specm1}, VG15, VG07\\
    \hline
    Measured spectrum below the prediction band but shape is similar & Gap was slightly larger than predicted or bonding was incomplete& VS21-F1, Figure \ref{VS21specf1}, VS09, VG15, \\
    \hline
    Measured spectrum above the prediction band but shape is similar & Gap was slightly smaller than predicted & VS21-M2\\
    \hline
    Measured spectrum exceeds DSP reference puck & unanticipated anti-reflection coating on the surface of the Si sample (e.g. Sharpie marker residue) & VG08\\
    \hline
    High frequency, low amplitude wiggles superimposed upon smooth underlying trend & Large, unanticipated gap, with a small fill factor  & VS21-C3\\   
     																			     & wavelength dependent diffraction from the mesh patterns & VS21-C3\\       
     																			     & measurement artifact  & VS21-C3\\       												
    \hline
    Non-unit transmission constant with wavelength (gray spectrum) & Artifact of measurement- vignetting of beam, absorptive coating, zero-point drift & VS21-C4 \\
    \hline
    Completely different spectrum than anticipated with low absolute transmission & Large, inhomogeneous gap consistent with an IR bubble & VG09 \\
    \hline
    \end{tabular}
  \end{center}
\end{table}

@@

\subsection{Mitigation of measurement artifacts}
The principle artifacts and their interpretation are listed in Table \ref{tabInterpretation}.  We mitigated the artifacts in the following ways.  For the Sharpie marker behaving as an anti-reflection (AR) coating, we simply removed the sharpie with IPA.  We have mostly been able to rule out wavelength-dependent diffraction from the mesh pattern as a measurement artifact, since the beam size is large and the angular spread of diffraction from the mesh pattern would be much smaller than the spread of the beam.  Still, to remove this concern, we moved to using the coarse mesh rather than the medium or fine meshes, when possible.  We saw no difference in the bond quality surrounding the medium or fine meshes, which is consistent with results from the literature \cite{1992JEMat..21..669M}.  Specifically, the gas pressure of bonded wafers grows as the fill factor of voids decreased.  In other words, what matters is the fill factor of voids, not their relative sizes.  Since the fine, medium, and coarse meshes had the same fill factors, we anticipate no difference in the pressure of the voids, and therefore no difference in the bonding properties.

@@

The consequence of our immersion grating design is a non-normal incidence interface in double pass.  In this case, the Fabry-P\'{e}rot framework developed here fails to predict the spectrum of the interface loss attributable to a finite albeit small (few nm) gap.  Specifically, complete internal reflection and evanescent waves will matter at large incidence angles, and the incoherent matrix technique is no longer accurate.  Polarization effects are important.  We do not have a mechanism for evaluating the bond performance until the final device is completed and its wavelength-dependent transmission is measured.  We can detect the performance degradation on final devices in the following nuanced way.  We will measure the efficiency of the bonded Si immersion grating as a function of both wavelength and angle.  Our group built a custom tool to make these measurements- the Custom Robotic Order, Wavelength, and Blaze Angle Recorder (CROWBAR) \cite{2012SPIE.8450E..2SG}.  Figure 3 in \cite{2012SPIE.8450E..2SG} shows the directly measured spectrum of a monolithic immersion grating.  Notably, the spectrum in Figure 3 in \cite{2012SPIE.8450E..2SG}  shows dozens of blaze peaks where the efficiency should be maximized.  A perfectly contacted Si$-$Si direct bonded optic would have identical performance to a monolithic counterpart.  Bonded immersion gratings with finite gap sizes will superimpose a wavelength dependent transmission loss on top of the blaze peaks.  In this way, the presence or absence of deleterious gaps can be identified in final immersion gratings.  One drawback of this approach is that gaps can only be identified after the costly process of producing the final optical device.  Large gaps identified this late in the game could render the costly optic useless, both in terms of efficiency and wavefront performance.  Our group is exploring the effectiveness of high-temperature annealing to eliminate interfacial gaps in completed devices.
